\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor=blue, linkcolor=black}
\usepackage{array}
\usepackage{xcolor}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}

\title{\textbf{Yaz: an editable, auditable tiny knowledge model\\ with routing-confidence abstention}}
\author{Tilelli LAB \\ \texttt{hello@tilelli.tech} \\ Marrakech}
\date{Technical report, v0.0.1 --- 17 June 2026}

\begin{document}
\maketitle

\begin{abstract}
We present \textbf{Yaz}, a sub-1M-parameter, byte-level language model designed so that individual
facts can be created, read, updated, and deleted (CRUD) one at a time, with \textbf{provable per-edit
locality}, and so that the model \textbf{abstains} when it is not confident which fact a prompt refers
to. Each fact is stored in its own addressable decoder column (an \emph{atom}); prompts are routed to
atoms by meaning using a frozen sentence embedding, so paraphrases reach the same fact. We report what
is verifiable on a small synthetic benchmark (50 country$\rightarrow$capital facts): exact
in-distribution edit reliability, zero-collateral deletion, structural locality across 40 sequential
edits (bits-per-character change $+0.000\%$), paraphrase-robust routing (held-out reach 0.696 vs 0.216
for surface routing), and a near-oracle abstention signal (risk-coverage AURC 0.004 against an oracle
bound of 0.003). We are equally explicit about what Yaz does \textbf{not} do: it is a \emph{first-byte}
editor (multi-byte generation is not faithful; full-word transfer $\approx 0.05$), routing degrades on
oblique name-free clues, and the scope is tiny and synthetic. Yaz is a clean, reproducible prototype in
the established model-editing lineage (ROME/MEMIT, GRACE, SERAC, PENME), \textbf{not} a state-of-the-art
result and not a defensible new capability. Its contribution is engineering clarity.
\end{abstract}

\section{Motivation}
Large models store facts diffusely across billions of weights. Two consequences follow: editing one
fact risks perturbing others, and a confidently wrong answer is indistinguishable from a correct one.
Yaz is a deliberately small experiment in the opposite design: put each fact in its own slot, make
edits surgical and reversible, and let the model decline when it is unsure which fact is being asked
about. The goal is \emph{auditability}, not capability.

\section{Method}
\textbf{Fact-atoms.} Yaz is a byte-level transformer whose output head is augmented with an additive
\emph{FactAtomLayer}: fact $i$ owns decoder column $i$. When a prompt is routed to fact $i$, that
column's contribution biases the first generated byte toward the stored answer.

\textbf{Routing by meaning.} A frozen sentence embedder
(\texttt{sentence-transformers/all-MiniLM-L6-v2}, 384-d, L2-normalized) maps a prompt to a vector.
Each fact's \emph{centroid} is the mean embedding of its training-template phrasings; a prompt routes
to the nearest centroid by cosine similarity. The embedder is never fine-tuned on the facts, and
centroids are built only from training templates --- held-out paraphrases never enter fitting, so
routing generalization is measured leak-free.

\textbf{CRUD verbs.} UPDATE swaps one decoder column; DELETE zeroes an atom; CREATE allocates a fresh
atom; READ is just routing. None require gradient steps.

\textbf{Abstention.} The routing \emph{confidence margin} (top-1 minus top-2 centroid cosine) is used
as a selective-prediction signal: below a threshold the model refuses rather than emitting a guess.

\section{Results (verified, reproducible)}
Benchmark: 50 country$\rightarrow$capital facts, single seed (2026), CPU. All numbers reproduce exactly
with the public sentence-transformers embedder (no internal dependencies).

\begin{center}
\begin{tabular}{L{6.2cm} L{8.0cm}}
\hline
\textbf{Capability} & \textbf{Measured result} \\
\hline
UPDATE (edit, no retraining) & in-distribution reliability 1.000; edits land 8/8 (first byte) \\
DELETE & fact removed, 0 collateral on other facts \\
CREATE & passes 4/4-condition battery (monosemantic / local / readable / deletable) \\
Per-edit locality & 0/10 collateral; bpc change $+0.000\%$ across 40 sequential edits \\
Sequential robustness & retention flat (1.000) over 40 edits \\
Paraphrase-robust routing & held-out reach \textbf{0.696} vs 0.216 surface (lexical) routing \\
Abstention (all, $n{=}298$) & risk-coverage \textbf{AURC 0.004} (oracle 0.003; grace 0.052; random 0.087) \\
Abstention operating point & coverage 0.966 at $\le 5\%$ answered-risk (vs 0.581 grace-distance) \\
Abstention (hard-only, $n{=}48$) & AURC 0.194 (oracle 0.159; grace 0.436; random 0.591) \\
\hline
\end{tabular}
\end{center}

\section{Limitations and honest scope}
These are not footnotes; they bound every claim above.
\begin{itemize}
\item \textbf{First-byte editor.} Edits reliably set the answer's \emph{first byte}. Multi-byte
generation is not faithful --- full-word transfer is $\approx 0.05$.
\item \textbf{A retracted claim.} An earlier ``edit-generalization'' headline of 0.675 was
\textbf{retracted}: a random-column-swap control produces a null of $\approx 0.688$, so that first-byte
number sits \emph{at chance}. What survives is routing \emph{reach}, not an edit-magnitude effect.
\item \textbf{Fragile routing.} Reach is $\approx 0.85$ on famous entities but falls to $\approx 0.50$
on oblique, name-free clues; answered coverage drops accordingly.
\item \textbf{Structural, not unconditional, locality.} The guarantee holds while no two facts route to
the same atom; collisions can appear at larger fact counts.
\item \textbf{Tiny, synthetic scope.} 50 facts, one seed, CPU. Not validated at scale or on
open-vocabulary knowledge.
\item \textbf{Not a moat.} The mechanisms each exist in the literature ($\S$6). Yaz combines them
cleanly and reproducibly; that is engineering, not a defensible capability.
\end{itemize}

\section{Reproducibility}
\begin{verbatim}
pip install -r requirements.txt
python scripts/scaling/s3_route_abstain.py     # abstention risk-coverage (AURC)
python scripts/scaling/s4_create_primitive.py  # the CREATE 4-condition battery
pytest -q                                       # in-dist reliability 1.000 + delete locality
\end{verbatim}
All runs are deterministic (seed 2026), CPU-only. The routing/abstention numbers were regenerated
end-to-end with the public \texttt{all-MiniLM-L6-v2} embedder, with no local paths or private packages,
and matched the reported values exactly.

\section{Related work}
Yaz sits in the knowledge-/model-editing lineage: locate-and-edit methods (ROME, MEMIT),
codebook/adapter editors (GRACE), retrieval-routed editors (SERAC), and embedding-keyed editing (PENME,
arXiv:2408.10411), with EasyEdit consolidating the family. The abstention component draws on selective
prediction and risk-coverage analysis. Yaz's only novel framing is the integration: routing-confidence
used directly as a refusal signal inside a CRUD editor with structural locality.

\section{Conclusion}
Yaz is an honest, small, reproducible demonstration that fact-level CRUD editing, structural locality,
and confidence-based abstention can coexist in a sub-1M-parameter byte-level model. It is a step, not a
breakthrough. We release it as a dated, open, reproducible artifact.

\vspace{1em}
\hrule
\vspace{0.5em}
{\small License: MIT. Code: \url{https://github.com/TilelliLab/yaz} \quad Weights:
\url{https://huggingface.co/TilelliLab/yaz}}

\end{document}
